The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-31 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Randall's comments

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.

@@

In this article, we will look specifically at the RFI situation of the MWA \citep{mwa}. The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison region in Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be build near the location of the MWA.

@@

@@

To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying field of views. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Not all observations cover the DVB frequencies. }\label{tbl:obs-list}
\begin{center}%
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 30\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{3\%}\\
\hline
\multicolumn{4}{|r|}{\textbf{With uniform channel coverage}} & \textbf{1.13\%} & \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature. This shape is e.g. visible in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is effected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.

@@

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

@@

\section{Comparison with LOFAR} \label{ch:lofar-comparison}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown in O2012, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.

@@

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.

SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?. Second, \textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA. Finally, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.

>>>>

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based or satellite transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.

@@

In this article, we will look specifically at the RFI situation of the MWA \citep{mwa}. The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Obervatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA. TODO mention something about the extent to which protections have been put in place for radio quietness at the MRO. Randall: ``I'm sure there's a document somewhere that describes it, otherwise you could contact Carol Wilson at CSIRO who's responsible for such things. (carol.wilson@csiro.au)''

@@

Recently, experiments to use the moon as a calibrator for EoR experiments have shown that the reflection of the moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}. However, reflections are spatially isolated to the centre of the observed disk caused by the moon, and the high resolution of LOFAR allows separating the reflected and intrinsic power from the moon \citep{vedantham-2014-todo}. In another study it is described that objects in space of about a meter in size, such as satellites and space debris, also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}. While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.

@@

To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying fields of view. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Not all observations cover the DVB frequencies. (TODO this table excludes RF11 and RF12 DVB interference; needs fix)}\label{tbl:obs-list}
\begin{center}%
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 30\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{3\%}\\
\hline
\multicolumn{4}{|r|}{\textbf{With uniform channel coverage}} & \textbf{1.13\%} & \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature. This shape is e.g. visible in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.

@@

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

@@

\section{Comparison with LOFAR} \label{ch:lofar-comparison}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency subrange of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown in O2012, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low elevations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.

@@

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DVB bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVB signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.

SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA. Finally, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.
